The role of standing variation in geographic convergent adaptation

### 2015-02-15 standing-variation-paper.tex
making changes in response to reviewer and molly comments

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{green}(#1)} }
\newcommand{\gc2}[1]{{\it\color{purple}(#1)} }

\newcommand{\plr}[1]{{\it\color{blue}(#1)}}
\usepackage{lineno}  %%%% FOR AMNAT
\newcommand{\var}{\mathop{\mbox{Var}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

@@

\section*{Abstract}
The extent to which populations experiencing shared selective pressures
adapt through a shared genetic response
is relevant to many questions in evolutionary biology. 
In a number of well studied traits and species, 
it appears that convergent evolution within species is common. 
In this paper, 
we explore how standing, deleterious genetic variation contributes to
convergent genetic responses in a geographically spread population,
extending our previous work on the topic.
Geographically limited dispersal slows the spread of each selected allele,
hence allowing other alleles 
-- newly arisen mutants or present as standing variation -- 
to spread before any one comes to dominate the population. 
When such alleles meet, their progress is substantially slowed 
-- if the alleles are selectively equivalent, 
they mix slowly, dividing the species range into a random tessellation,
which can be well understood by analogy to a Poisson process model of crystallization. 
In this framework, 
we derive the geographic scale over which a typical allele is expected to dominate, 
the time it takes the species to adapt as a whole,
and the proportion of adaptive alleles that arise from standing variation.
Finally, we explore how negative pleiotropic effects of alleles before an environment change 
can bias the subset of alleles that get to contribute to a species adaptive response. 
We apply the results to the many geographically localized G6PD
deficiency alleles thought to confer resistance to malaria, \gc2{where the large
mutational target size of G6PD makes likely candidate for adaptation
from standing variation, despite the selective cost of G6PD
deficiency.}
We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
\gc2{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
Our results suggest that much of adaptation may be geographically local 
even when selection pressures are widespread. \gc2{Therefore, we argue
that caution must be exercised when arguing that strongly
geographically restricted alleles are necessarily the outcome of local adaptation.}
We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.

@@

There are an increasing number of examples where
different populations within a species have
adapted to similar environments by means of independent genetic changes. 
In some cases this convergent evolution \gc2{is the result of quite
distinct genetic changes, involving very different genes and pathways} despite shared selection pressures; 
in other cases independent adaptations are identical down the same
nucleotide change
\citep{Jeong20141,stern2013genetic,MartinOrgogozo:13, Conte2012}. 
Such convergent evolution within populations has been seen for many
carefully studied phenotypes across a range of species, 
including drug resistance in pathogens,
resistance to pathogens or pesticides,
and the molecular basis of pigmentation changes.
The phrase ``parallel evolution'' is also used to refer to such convergent evolution;
here we use these synonymously, 
as we are concerned with adaptation within a single
species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. 
These include the extent to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
and to \gc2{what degree species should be regarded} as cohesive units. 
Convergent evolution also affects our ability to detect adaptation from population genomic data,
since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.

@@

Spatial population structure,
as caused for example by geographically limited dispersal, 
also increases the chance of convergent evolution. 
For example, 
\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
since alleles are unable to spread through intervening populations
\citep{RalphCoop:14}. 
In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
 a spatially spread population, where dispersal is local,
that is exposed to some novel, spatially homogeneous selection pressure.
We assumed that a single mutational change was sufficient to adapt
the population after the change in environment. Under this assumption,
selected alleles arise and spread locally. However, if the geographic
area is large enough multiple selected alleles can
arise independently and spread before any one has spread across all of
space ( Figure \ref{fig:sim-spread}).  one allele's rapid spatial spread can be
halted by running into convergent allele. See Figure
\ref{fig:sim-spread}. A somewhat analogous situation also arises in spatial models of clonal
inference in asexuals \citep{martens_interfering_2011}.}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
  \end{center}
  \caption{
{\bf Simulation of the geographic spread of new mutations}. Simulation on a grid of $101 \times 101$ populations with 100 individuals in each, showing spread of a beneficial allele with advantage $s=0.1$.  Times are in numbers of generations, with probability of mutation per generation 0.3. 
}
  \label{fig:sim-spread}
\end{figure}

\gc{Under this model we previously derived} the characteristic geographic scale over which 
multiple instances of the adaptive allele are expected to arise in parallel. 
This characteristic length could be expressed in terms of a simple compound
parameter \gc2{comprised of} our parameters of interest. \gc{In
\citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
allele (e.g. because the allele was very strongly deleterious before
the environmental change)}, 
so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.

In this paper, we extend this spatial model to include standing variation 
present at mutation-selection balance before the selection pressures switch.
Below,
we show that convergent adaptation within a widespread species is likely to be
common \gc{when ranges, population sizes, or mutational targets are large}, as has already been seen for a number of traits. 
On this basis owe argue 
that the genetics of adaptation may often be geographically local \gc{even
when selection pressures are geographically broad}, 
% and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
and that widespread selective sweeps \gc2{should tend to} occur
when adaptation is highly constrained (e.g.\ by small mutation rate or
the need for a linked combination of alleles).
We discuss the \gc{history, and implications,} of this view for the evolutionary
coherence of species and molecular evolution.

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
homozygote for the mutant alleles. \gc2{Note that this means that we
  are ignoring the case of totally recessive alleles.}
\gc{We assume that the variance in offspring number is one, see
  \cite{ralphcoop2010} for how this assumption can be relaxed.}
%The number of offspring has finite variance, 
%which would divide the probability of establishment of advantageous alleles below,
%but to keep the formulas simpler we assume to be equal to one.
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared geographic distance between parent and child
is $\sigma^2$. \gc2{Note that we assume that the dispersal kernel is
  not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
  discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
  on rapid spreading due to very long distance dispersal.}
The species occupies an area with mean density
$\rho$ of haploid individuals \gc{(twice the number of diploids)} per unit area.

@@

Before $t=0$, on the other hand, the allele is deleterious.
The genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for \gc2{a short} time (note that we have assumed that $s_d$ is not too small).
\gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
well--approximated by a 
Poisson process with rate $\mu \rho$ \gc{(if $\mu \rho \ll 1$) as
mutations are rare events. Therefore,} 
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define
to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
\gc2{Our $\lambda_0$ is the mean density
of standing variants that are present and escape loss when the
environment shifts.}
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, 
we can then treat each such family as equivalent to a single new mutation,
but with somewhat larger probability of local establishment,
\gc{following the environmental shift}.
(This approximation will be good 
if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let \gc{$Z_T$ be the number of its descendants at time $0$}.
At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s$ of establishing, approximately independently.
Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the mean number of clusters of standing variants destined to fix locally, per unit area, is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
        & =  \frac{\lambda }{ (-log(1-s_d)) }.\\
\end{align}
Note that this is closely related to the equilibrium frequency of an
allele at mutation selection balance $\mu/s_d$
\citep{Haldane:27,Haldane:37, Orr:01}.

@@

Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
and spread through human populations in areas where malaria \gc2{has been} endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes that confer malaria resistance have been observed in the $\beta$-globin gene;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutation at different locations within Africa
\citep{Flint:98,ralphcoop2010}.
Another particularly impressive case of convergent evolution \gc2{is} the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Whether these all confer resistance to malaria is unknown,
but malaria is thought to be the primary driver of these polymorphisms
\citep[see][ for a general review]{hedrick2011population}.
Three G6PD deficiency alleles are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
\gc2{ 
The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
across study populations and the form of malaria considered
\citep[see ][ for recent discussion]{manjurano_2015, malaria_network_2014}.  
The $A-$ and Mahidol alleles are thought to offer some protective effects against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
  Louicharoen-g6pd, manjurano_2015}. }
Haplotype-based analysis of genetic diversity surrounding
$A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that these three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$
\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
This is in reasonable agreement with the selection coefficients calculated by
\citet{Ruwende-g6pd} on the basis of the present day levels of
resistance to malaria due to the $A-$ allele. \\

@@

Given \gc2{such strong selection the alleles} should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
and homozygote females may also not be protected against malaria \citep{manjurano_2015, malaria_network_2014}. 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
\geq 0.05$),
in order to have resulted in the equilibrium frequency seen today in
areas with endemic malaria \citep[based on heterozygote advantage
calculations for the X chromosome, results not shown, see also][]{Ruwende-g6pd}.

@@

In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb}, we show the range of characteristic lengths as a function 
of various parameters chosen to match the evolution of malaria
resistance at G6PD. 
These curves match our intuition that \gc{higher} population densities result in
smaller characteristic lengths (as would higher mutation rates). 
Allowing standing variation increases \gc{the 
input of new alleles} and so decreases the characteristic length below that predicted
by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
In turn, increasing the prior deleterious effect of the allele ($s_d$)
acts to increase the
characteristic length until it reaches that predicted by new mutation alone. 
Higher dispersal distances lead to larger characteristic lengths, 
since more rapid geographic spread block other mutations from establishing. 
A larger selective advantage ($s_b$) acts in two conflicting
ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
\gc2{to} escape drift and become established. 
The effect of helping\gc{locally} establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
 (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
as in that case the speed of spread does not matter \gc2{only the
  initial density of established alleles after the environmental shift}. 
The dependence of the characteristic length on $s_b$ when only new
mutations contribute is much weaker (\gc{of order} $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
Overall, the range of characteristic lengths observed are reasonably
consistent with the average diameter of a G6PD variant in Eurasia of
800km, especially for the lower population density, as long as the
fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \gc2{high}.

@@

In Figure \ref{G6PD_standing_var_proportion} we show the proportion of
alleles that spread from standing variation, and the proportion of
geographic space covered by standing variants for parameters chosen to match our G6PD example.
Even for relatively large deleterious costs prior to the environmental switch, 
standing variants still makes up quite a large proportion of the adaptive alleles, 
and an even larger proportion of the range 
(they \gc2{occupy a larger area than new mutations, since they get a
headstart}).

@@

Each of these local sweeps will be associated with the haplotype on which the particular allele arose. 
Under the parameter regime we study,
standing variants are still quite young, 
so we do not expect a strongly reduced hitchhiking effect. 
As such, following the initial period of
adaptation, we should expect the population to be partitioned into a
set of geographically restricted long haplotypes. 
Given sufficient time these haplotypes will mix together through
migration and drift, 
potentially leading to a \gc2{within population} signal of a sweep from multiple independent
mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
or to multiple partial sweeps if the loci are scattered across the
genome \citep{CoopRalph:12}.

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{blue} #1} }
\newcommand{\mfp}[1]{{\it\color{red} #1} }

\newcommand{\plr}[1]{{\it\color{green}(#1)}}
\usepackage{lineno}  %%%% FOR AMNAT
\newcommand{\var}{\mathop{\mbox{Var}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

@@

\section*{Abstract}
The extent to which populations experiencing shared selective pressures
adapt through a shared genetic response
is relevant to many questions in evolutionary biology. 
In a number of well studied traits and species, 
it appears that convergent evolution within species is common. 
In this paper, 
we explore how standing, deleterious genetic variation contributes to
convergent genetic responses in a geographically spread population,
extending our previous work on the topic.
Geographically limited dispersal slows the spread of each selected allele,
hence allowing other alleles 
-- newly arisen mutants or present as standing variation -- 
to spread before any one comes to dominate the population. 
When such alleles meet, their progress is substantially slowed 
-- if the alleles are selectively equivalent, 
they mix slowly, dividing the species range into a random tessellation,
which can be well understood by analogy to a Poisson process model of crystallization. 
In this framework, 
we derive the geographic scale over which a typical allele is expected to dominate, 
the time it takes the species to adapt as a whole,
and the proportion of adaptive alleles that arise from standing variation.
Finally, we explore how negative pleiotropic effects of alleles before an environment change 
can bias the subset of alleles that get to contribute to a species adaptive response. 
We apply the results to the many geographically localized G6PD
deficiency alleles thought to confer resistance to malaria, \mfp{where the large
mutational target size makes it a likely candidate for adaptation
from standing variation, despite the selective cost of G6PD
deficiency alleles in the absence of malaria.}
We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
\mfp{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
Our results suggest that much of adaptation may be geographically local 
even when selection pressures are widespread. \mfp{Therefore, we argue
that caution must be exercised when arguing that strongly
geographically restricted alleles are necessarily the outcome of local adaptation.}
We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.

@@

There are an increasing number of examples where
different populations within a species have
adapted to similar environments by means of independent genetic changes. 
In some cases this convergent evolution \mfp{is the result of quite
distinct genetic changes, involving very different genes and pathways} despite shared selection pressures; 
in other cases independent adaptations are identical down the same
nucleotide change
\citep{Jeong20141,stern2013genetic,MartinOrgogozo:13, Conte2012}. 
Such convergent evolution within populations has been seen for many
carefully studied phenotypes across a range of species, 
including drug resistance in pathogens,
resistance to pathogens or pesticides,
and the molecular basis of pigmentation changes.
The phrase ``parallel evolution'' is also used to refer to such convergent evolution;
here we use these synonymously, 
as we are concerned with adaptation within a single
species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. 
These include the extent to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
and to \mfp{what degree species should be regarded} as cohesive units. 
Convergent evolution also affects our ability to detect adaptation from population genomic data,
since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.

@@

Spatial population structure,
as caused for example by geographically limited dispersal, 
also increases the chance of convergent evolution. 
For example, 
\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
since alleles are unable to spread through intervening populations
\citep{RalphCoop:14}. 
In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
 a spatially spread population, where dispersal is local,
that is exposed to some novel, spatially homogeneous selection pressure.
We assumed that a single mutational change was sufficient to adapt
the population after the change in environment. Under this assumption,
selected alleles arise and spread locally. However, if the geographic
area is large enough multiple selected alleles can
arise independently and spread before any one has spread across all of
space ( Figure \ref{fig:sim-spread}).  one allele's rapid spatial spread can be
halted by running into convergent allele. See Figure
\ref{fig:sim-spread}. A somewhat analogous situation also arises in spatial models of clonal
inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
  \end{center}
  \caption{
{\bf Simulation of the geographic spread of new mutations}. Simulation
on a grid of $101 \times 101$ populations with 100 haploid individuals
in each, showing spread of a beneficial allele with advantage $s=0.1$.
Times are in numbers of generations, with probability of mutation per
generation 0.3. Different colours show the different independent
origins and spreads of selected alleles. Alleles quick establish and
start to spread, but because they are selective equivalent once they
spread into each other they only mix slowly (note the lack of change
from generation 600 onward). More details of the
simulation are given in \citet{ralphcoop2010}. 
}
  \label{fig:sim-spread}
\end{figure}

\gc{Under this model we previously derived} the characteristic geographic scale over which 
multiple instances of the adaptive allele are expected to arise in parallel. 
This characteristic length could be expressed in terms of a simple compound
parameter \mfp{comprised of} our parameters of interest. \gc{In
\citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
allele (e.g. because the allele was very strongly deleterious before
the environmental change)}, 
so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.

In this paper, we extend this spatial model to include standing variation 
present at mutation-selection balance before the selection pressures switch.
Below,
we show that convergent adaptation within a widespread species is likely to be
common \gc{when ranges, population sizes, or mutational targets are large}, as has already been seen for a number of traits. 
On this basis owe argue 
that the genetics of adaptation may often be geographically local \gc{even
when selection pressures are geographically broad}, 
% and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
and that widespread selective sweeps \mfp{should tend to} occur
when adaptation is highly constrained (e.g.\ by small mutation rate or
the need for a linked combination of alleles).
We discuss the \gc{history, and implications,} of this view for the evolutionary
coherence of species and molecular evolution.

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
homozygote for the mutant alleles. \mfp{Note that this means that we
  are ignoring the case of totally recessive alleles.}
\gc{We assume that the variance in offspring number is one, see
  \cite{ralphcoop2010} for how this assumption can be relaxed.}
%The number of offspring has finite variance, 
%which would divide the probability of establishment of advantageous alleles below,
%but to keep the formulas simpler we assume to be equal to one.
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared geographic distance between parent and child
is $\sigma^2$. \mfp{Note that we assume that the dispersal kernel is
  not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
  discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
  on rapid spreading due to very long distance dispersal.}
The species occupies an area with mean density
$\rho$ of haploid individuals \gc{(twice the number of diploids)} per unit area.

@@

Before $t=0$, on the other hand, the allele is deleterious.
The genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
\gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
well--approximated by a 
Poisson process with rate $\mu \rho$ \gc{(if $\mu \rho \ll 1$) as
mutations are rare events. Therefore,} 
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define
to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
\mfp{Our $\lambda_0$ is the mean density
of standing variants that are present and escape loss when the
environment shifts.}
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, 
we can then treat each such family as equivalent to a single new mutation,
but with somewhat larger probability of local establishment,
\gc{following the environmental shift}.
(This approximation will be good 
if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let \gc{$Z_T$ be the number of its descendants at time $0$}.
At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s$ of establishing, approximately independently.
Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the mean number of clusters of standing variants destined to fix locally, per unit area, is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
        & =  \frac{\lambda }{ (-log(1-s_d)) }.\\
\end{align}
Note that this is closely related to the equilibrium frequency of an
allele at mutation selection balance $\mu/s_d$
\citep{Haldane:27,Haldane:37, Orr:01}.

@@

Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
and spread through human populations in areas where malaria \mfp{has been} endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes that confer malaria resistance have been observed in the $\beta$-globin gene;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutation at different locations within Africa
\citep{Flint:98,ralphcoop2010}.
Another particularly impressive case of convergent evolution \mfp{is} the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Whether these all confer resistance to malaria is unknown,
but malaria is thought to be the primary driver of these polymorphisms
\citep[see][ for a general review]{hedrick2011population}.
Three G6PD deficiency alleles are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
\mfp{ 
The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
across study populations and the form of malaria considered
\citep[see ][ for recent discussion]{manjurano_2015, malaria_network_2014}.  
The $A-$ and Mahidol alleles are thought to offer some protective effects against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
  Louicharoen-g6pd, manjurano_2015}. }
Haplotype-based analysis of genetic diversity surrounding
$A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that these three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$
\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
This is in reasonable agreement with the selection coefficients calculated by
\citet{Ruwende-g6pd} on the basis of the present day levels of
resistance to malaria due to the $A-$ allele. \\

@@

Given \mfp{such strong selection the alleles} should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
and homozygote females may also not be protected against malaria \citep{manjurano_2015, malaria_network_2014}. 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
\geq 0.05$),
in order to have resulted in the equilibrium frequency seen today in
areas with endemic malaria \citep[based on heterozygote advantage
calculations for the X chromosome, results not shown, see also][]{Ruwende-g6pd}.

@@

In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb}, we show the range of characteristic lengths as a function 
of various parameters chosen to match the evolution of malaria
resistance at G6PD. 
These curves match our intuition that \gc{higher} population densities result in
smaller characteristic lengths (as would higher mutation rates). 
Allowing standing variation increases \gc{the 
input of new alleles} and so decreases the characteristic length below that predicted
by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
In turn, increasing the prior deleterious effect of the allele ($s_d$)
acts to increase the
characteristic length until it reaches that predicted by new mutation alone. 
Higher dispersal distances lead to larger characteristic lengths, 
since more rapid geographic spread block other mutations from establishing. 
A larger selective advantage ($s_b$) acts in two conflicting
ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
\mfp{to} escape drift and become established. 
The effect of helping\gc{locally} establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
 (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
as in that case the speed of spread does not matter \mfp{only the
  initial density of established alleles after the environmental shift}. 
The dependence of the characteristic length on $s_b$ when only new
mutations contribute is much weaker (\gc{of order} $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
Overall, the range of characteristic lengths observed are reasonably
consistent with the average diameter of a G6PD variant in Eurasia of
800km, especially for the lower population density, as long as the
fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \mfp{high}.

@@

In Figure \ref{G6PD_standing_var_proportion} we show the proportion of
alleles that spread from standing variation, and the proportion of
geographic space covered by standing variants for parameters chosen to match our G6PD example.
Even for relatively large deleterious costs prior to the environmental switch, 
standing variants still makes up quite a large proportion of the adaptive alleles, 
and an even larger proportion of the range 
(they \mfp{occupy a larger area than new mutations, since they get a
headstart}).

@@

Each of these local sweeps will be associated with the haplotype on which the particular allele arose. 
Under the parameter regime we study,
standing variants are still quite young, 
so we do not expect a strongly reduced hitchhiking effect. 
As such, following the initial period of
adaptation, we should expect the population to be partitioned into a
set of geographically restricted long haplotypes. 
Given sufficient time these haplotypes will mix together through
migration and drift, 
potentially leading to a \mfp{within population} signal of a sweep from multiple independent
mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
or to multiple partial sweeps if the loci are scattered across the
genome \citep{CoopRalph:12}.
